WED-SQL (Draft)

### 2016-05-19 doc/paper/main_en.tex
paper: ch3

% *** SPECIALIZED LIST PACKAGES ***
%
\usepackage{algorithmic}
\usepackage{algorithm}
\floatname{algorithm}{Algoritmo}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

@@

try:
    w.run()
except KeyboardInterrupt:
    print()
    sys.exit(0)
\end{Verbatim}
\caption{Standard WED-worker implementation in Python}
\label{fig_ww}
\end{figure}

The first step to write a WED-worker using the BaseWorker package is to import the module \emph{BaseClass}, which in fact
is an abstract class that implements the communication protocol with the WED-server as well as manages all connections
and transactions involved.  
\\
\indent  Next, a concrete class must implement the abstract method \emph{wed\_trans()} from BaseClass and initialize the following
class attributes: 
\begin{itemize}
\item \emph{trname}: WED-transition name that will be carried out by this WED-worker. Must match the name used on the WED-flow definition;
\item \emph{dbs}: WED-server connection parameters conforming the \emph{psycopg2} driver format. At least the username, database name and
                 the WED-worker name must be provided;
\item \emph{wakeup\_interval}: Time interval in which the WED-worker execution is suspended (sleeping) if there is no new WED-transitions
                              to be performed (explained in details on section~\ref{not}); 
\end{itemize}

@@

\section{Algoritmos}
Após definir e carregar um WED-flow em uma base de dados do WED-server e inicializar seus respectivos WED-workers, uma
nova instância é inicializada inserindo-se um WED-state na tabela WED\_flow. Essa nova instância receberá um identificador
único que será utilizado tanto para o registro de sua história de execução, na tabela WED\_trace, quanto para o controle
transacional das WED-transitions. Se os valores padrão definidos para os WED-attributes, por meio da coluna "adv", na 
tabela WED\_attr forem os valores de um WED-state inicial, basta executar o comando abaixo para se inicializar uma nova
instância:
\begin{Verbatim}[fontsize=\small]

     INSERT INTO wed_flow DEFAULT VALUES;
\end{Verbatim}
  Para cada nova instância inserida na tabela wed\_flow, o WED-server irá comparar os predicados das WED-conditions definidos
na tabela wed\_trig com o WED-state representado na nova instância. Para cada condição satisfeita a respectiva WED-transition
será disparada na forma de uma entrada na tabela job\_pool e uma notificação será enviada ao WED-worker responsável
por executá-la. Além disso, serão adicionadas novas entradas na tabela wed\_trace registrando os eventos ocorridos. O WED-server
ficará então aguardando até que uma nova instância ou uma WED-transition seja iniciada. Vale notar que WED-states iniciais 
que não sejam finais e que não disparem ao menos uma WED-transition serão rejeitados pelo WED-server.
\indent  Para iniciar uma WED-transition, os WED-workers podem atuar de dois modos distintos: imediatamente ao receber 
uma notificação do WED-server, ou quando o limite de tempo de espera por notificações termina (atributo wakeup\_interval) 
, nesse caso será feita uma consulta à tabela job\_pool em busca de WED-transitions pendentes. Independentemente do modo 
de atuação e antes de inicializar a transação, cada WED-worker precisa notificar o WED-server qual WED-transition e em qual 
instância do WED-flow será executada a transação. Essa notificação é feita requisitando-se uma trava especial chamada 
\emph{Advisory Lock}, que será discutida em detalhes na sesão V.
Essa trava é necessária tanto para avisar a outros WED-workers que estejam trabalhando na mesma WED-transition que não a 
executem para a mesma instância, quanto para que o WED-server possa controlar o tempo limite da execução de cada WED-transition. 
Não é possivel executar uma WED-transition sem que o respectivo WED-worker consiga obter essa trava previamente.
\indent De posse da referida trava, um WED-worker terá que finalizar a transação dentro do tempo limite de execução da WED-transition.
Essa transação é finalizada atualizando-se o WED-state atual da instância em que está sendo executada a WED-transition por meio
de um \emph{UPDATE} na tabela wed\_flow com o valor retornado pelo método wed\_trans().
\indent Quando uma instância é atualizada, o WED-server novamente irá verificar se esse novo WED-state dispara novas WED-transitions
e, em caso afirmativo, irá registrá-las na tabela job\_pool, além de atualizar o histórico de execução na tabela wed\_trace. 
Caso esse novo WED-state satisfaça a condição final e não haja nenhuma WED-transition pendente, essa instancia é marcada como finalizada e não poderá ser
modificada, a menos que sejam inseridos novos WED-attributes ou que sejam modificadas ou adicionadas novas WED-triggers. Em
caso de não haver WED-transitions pendentes e esse WED-state não for final, a instância será marcada como exceção e uma
entrada especial será adicionada a tabela job\_pool para indicar que algum mecanismo de recuperação deve ser iniciado.
\begin{algorithm}
\caption{WED-server}
\label{alg1}
\begin{algorithmic}[1]
\REQUIRE WED-flow instance $i$
\ENSURE \textbf{true} for a successfully transaction,\\
         \hspace{23pt}\textbf{false} on the otherwise
\STATE $L \leftarrow$ empty WED-transitions list 
\STATE $s \leftarrow$ current WED-state of $i$
\FOR{$tg$ in WED-triggers}
\STATE $c,tr \leftarrow$ $tg($WED-condition$)$,$tg($WED-transition$)$
\IF{$s$ satisfies $c$}
\IF{$c$ is the final condition}
\STATE insert $\_FINAL$ in $L$
\ELSE
\STATE insert $tr$ in $L$
\ENDIF
\ENDIF
\ENDFOR
\IF{$L$ is empty}
\IF{$i$ is a new instance}
\STATE Reject $i$ and abort the transaction
\RETURN \FALSE
\ELSIF{$i$ \textbf{DOES NOT} has any pending WED-transition}
\STATE Set $i$ as an exception
\STATE Insert \_EXCPT into Job\_Pool
\ENDIF
\STATE Update $i$ at WED\_trace
\RETURN \TRUE
\ELSIF{$L$ contains $\_FINAL$}
\IF{$i$ has pending WED-transitions}
\STATE Reject $i$ and abort the transaction
\RETURN \FALSE
\ELSE
\STATE Set $i$ as a final WED-state
\STATE Update $i$ at WED\_trace
\RETURN \TRUE
\ENDIF
\ELSE
\FOR{$tr$ in $L$}
\STATE Insert $tr$ into Job\_Pool
\STATE Notify the WED-worker(s) responsible for $tr$
\RETURN \TRUE
\ENDFOR
\ENDIF

@@

\begin{algorithm}
\caption{WED-worker}
\label{alg1}
\begin{algorithmic}[1]
\REQUIRE WED-transition's name: \emph{trname}
\ENSURE \textbf{true} for a successfully transaction,\\
         \hspace{23pt}\textbf{false} on the otherwise
\LOOP
\STATE $n \leftarrow$ Null value
\WHILE{$n$ is Null}
\STATE $n \leftarrow wait\_for\_notification(trname,wkup)$
\IF [No notification has arrived after $wkup$ seconds]{$n$ is Null}
\STATE $n \leftarrow $ Ask WED-server for any $trname$ pending WED-transition
\ENDIF
\ENDWHILE
\STATE Initiate the transaction
\IF {got\_lock($n$)} 
\STATE Run the WED-transition $trname$ for a given $i$ WED-flow instance
\STATE Commit the transaction
\RETURN \TRUE
\ELSE
\STATE Abort the transaction
\RETURN \FALSE
\ENDIF  
\ENDLOOP

@@

\section{Gerenciamento transacional}
\label{GT}
Cada instância de um WED-flow pode ser vista como uma Transação Longa (LLT) e, de acordo com ~\cite{SGD87},
suas WED-transitions podem ser vistas como passos Saga. Sendo assim, é função do WED-server garantir a integridade dos estados
de dados, manter a consistência transacional e garantir que todo WED-flow termine ou em um estado final ou em um estado
de exceção e, nesse caso, deve permitir que um passo compensatório possa ser executado. 
\par
Nas tópicos seguintes serão apresentados em detalhes os mecanismos de controle transacional empregados e o protocolo de
comunicação entre múltiplos WED-workers e um WED-server.

>>>>

% *** SPECIALIZED LIST PACKAGES ***
%
\usepackage{algorithmic}
\usepackage{algorithm}
%\floatname{algorithm}{Algoritmo}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

@@

try:
    w.run()
except KeyboardInterrupt:
    print()
    sys.exit(0)
\end{Verbatim}
\caption{Standard WED-worker implementation in Python}
\label{fig_ww}
\end{figure}
The first step to write a WED-worker using the BaseWorker package is to import the module \emph{BaseClass}, which in fact
is an abstract class that implements the communication protocol with the WED-server as well as manages all connections
and transactions involved.  
\\
\indent  Next, a concrete class must implement the abstract method \emph{wed\_trans()} from BaseClass and initialize the following
class attributes: 
\begin{itemize}
\item \emph{trname}: WED-transition name that will be carried out by this WED-worker. Must match the name used on the WED-flow definition;
\item \emph{dbs}: WED-server connection parameters conforming the \emph{psycopg2} driver format. At least the username, database name and
                 the WED-worker name must be provided;
\item \emph{wakeup\_interval}: Time interval in which the WED-worker execution is suspended (sleeping) if there is no new WED-transitions
                              to be performed (explained in details on section~\ref{not}); 
\end{itemize}

@@

\section{Algorithms}

After a WED-flow application is defined, loaded into the WED-server and the WED-workers are initialized and ready to run,
new instances are created by inserting a WED-state into the WED\_flow table. The WED-server will assign to a new instance 
a WED-flow unique identifier (wid), which in turn will be used to both track its execution history as well as for transaction
management purposes. If the default values set for the WED-attributes, in the WED\_attr table, happens to match the ones
of a initial WED-state, then a new instance can be created by simply running the following SQL statement on the WED-server:

\begin{Verbatim}[fontsize=\small]

     INSERT INTO wed_flow DEFAULT VALUES;
\end{Verbatim}

When a new instance is initiated or when a WED-worker successfully commits a new WED-state, the WED-server henceforth runs 
its main algorithm (Alg.~\ref{alg1}). It starts by evaluating all WED-conditions predicates against the new WED-state. Then, for
each satisfied predicate, the WED-transition associated with this WED-condition will be fired and registered on the Job\_Pool table
,a notification will also be sent to the respective WED-worker. Besides that, this events will be added to the instance's
tracing record on the WED\_trace table. If a WED-state satisfies the final condition and if there is no pending WED-transitions,
either running or awaiting in the Job\_Pool table, for a given instance, it is then marked as final and cannot be further
modified. In the case when there is no pending WED-transitions and the new WED-state is not final, the instance will be
marked as an exception and a special record will be inserted into Job\_Pool to signal some external recovery mechanism~\cite{ICWS12}.
It is worth mentioning that unless a initial WED-state is also final, it must fire 
at least one WED-transition, otherwise the WED-server will reject the new instance. 
\\
\indent  WED-workers spend most of their idle time suspended, waiting for a notification send by the WED-server when there is work 
to be done. After some time waiting, if not notification has arrived they wake up and ask the WED-server for work (this
behavior will be explained in section~\ref{not}). Either way, before initiate a WED-transition, the WED-worker need to 
"inform" the WED-server of in which instance and what WED-transition it is about to perform. This is done by requesting the
WED-server a special lock. If this lock is granted, then the WED-worker can proceed (See alg.~\ref{alg2})
\\
\indent This lock system is necessary once more than one WED-worker can be simultaneously activated for the same WED-transition.
Therefore, its main role is to avoid the same WED-transition from being performed for the same WED-flow instance at the
same time. Furthermore, it also helps the WED-server to manage transactions. It is wise to assume that no WED-transition
can be committed by a WED-worker not holding a lock.
\\
\indent Once in possession of a lock, the WED-worker must finish its job within the time limit defined for the WED-transition.
The transaction is successfully finalized updating the current WED-state for this instance on the WED\_flow table.

\begin{algorithm}
\caption{WED-server}
\label{alg1}
\begin{algorithmic}[1]
\REQUIRE WED-flow instance $i$
\ENSURE \textbf{true} for a successfully transaction,\\
         \hspace{23pt}\textbf{false} on the otherwise
\STATE $L \leftarrow$ empty WED-transitions list 
\STATE $s \leftarrow$ current WED-state of $i$
\FOR{$tg$ in WED-triggers}
\STATE $c,tr \leftarrow$ $tg($WED-condition$)$,$tg($WED-transition$)$
\IF{$s$ satisfies $c$}
\IF{$c$ is the final condition}
\STATE insert $\_FINAL$ in $L$
\ELSE
\STATE insert $tr$ in $L$
\ENDIF
\ENDIF
\ENDFOR
\IF{$L$ is empty}
\IF{$i$ is a new instance}
\STATE Reject $i$ and abort the transaction
\RETURN \FALSE
\ELSIF{$i$ \textbf{DOES NOT} has any pending WED-transition}
\STATE Set $i$ as an exception
\STATE Insert \_EXCPT into Job\_Pool
\ENDIF
\STATE Update $i$ at WED\_trace
\RETURN \TRUE
\ELSIF{$L$ contains $\_FINAL$}
\IF{$i$ has pending WED-transitions}
\STATE Reject $i$ and abort the transaction
\RETURN \FALSE
\ELSE
\STATE Set $i$ as a final WED-state
\STATE Update $i$ at WED\_trace
\RETURN \TRUE
\ENDIF
\ELSE
\FOR{$tr$ in $L$}
\STATE Insert $tr$ into Job\_Pool
\STATE Notify the WED-worker(s) responsible for $tr$
\RETURN \TRUE
\ENDFOR
\ENDIF

@@

\begin{algorithm}
\caption{WED-worker}
\label{alg2}
\begin{algorithmic}[1]
\REQUIRE WED-transition's name: \emph{trname}
\ENSURE \textbf{true} for a successfully transaction,\\
         \hspace{23pt}\textbf{false} on the otherwise
\LOOP
\STATE $n \leftarrow$ Null value
\WHILE{$n$ is Null}
\STATE $n \leftarrow wait\_for\_notification(trname,wkup)$
\IF [No notification has arrived after $wkup$ seconds]{$n$ is Null}
\STATE $n \leftarrow $ Ask WED-server for any $trname$ pending WED-transition
\ENDIF
\ENDWHILE
\STATE Initiate the transaction
\IF {got\_lock($n$)} 
\STATE Run the WED-transition $trname$ for a given $i$ WED-flow instance
\STATE Commit the transaction
\RETURN \TRUE
\ELSE
\STATE Abort the transaction
\RETURN \FALSE
\ENDIF  
\ENDLOOP

@@

\section{Transactional Management}
\label{GT}
Cada instância de um WED-flow pode ser vista como uma Transação Longa (LLT) e, de acordo com ~\cite{SGD87},
suas WED-transitions podem ser vistas como passos Saga. Sendo assim, é função do WED-server garantir a integridade dos estados
de dados, manter a consistência transacional e garantir que todo WED-flow termine ou em um estado final ou em um estado
de exceção e, nesse caso, deve permitir que um passo compensatório possa ser executado. 
\par
Nas tópicos seguintes serão apresentados em detalhes os mecanismos de controle transacional empregados e o protocolo de
comunicação entre múltiplos WED-workers e um WED-server.
